---
title: "Visualizing U.S. Crime and Law Enforcement in the 2010s"
author: "Aqib Mahfuz"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r echo=F, warning=F, message=F}
# Loading libraries for project

library(tidyverse)
library(ggplot2)

library(rjson)  # for reading JSON files from API and more
library(jsonlite)
library(RJSONIO)

library(usmap)  # for easily plotting map of United States
library(ggpubr) # for putting multiple plots into one figure
```

## Background
Among the several divisive issues facing the United States in the early 2020s, the topic of law enforcement and policing has stood out in terms of the tension it has created within the American people. On one hand (Defund the police), and on the other (double down on law and order). To clarify, this experiment will not explicitly revolve around the issue of police brutality (statistics which are, regrettably, underreported). Rather, the goal is to analyze trends in crime and law enformcement between 2010 and 2019. Which types of crimes are the most prevalent and on the rise this past decade? 
We aim to tackle three central questions:
1. What effect does 



## Hypotheses



## Procedure

```{r echo=F, warning=F, message=F}
# Setting up API calls
creds <- rjson::fromJSON(file = '../crime_data_api_credentials.json')
API_KEY <- creds$api_key
CDE_URL <- "https://api.usa.gov/crime/fbi/sapi"

#paste(CDE_URL, API_KEY, sep="")

temp <- jsonlite::fromJSON("https://api.usa.gov/crime/fbi/sapi/api/summarized/arson/national/2010/2019?API_KEY=iiHnOKfno2Mgkt5AynpvPpUQTEyxE77jo1RU8PIv")

```

```{r echo=F, warning=F, message=F}
# Collecting lookup data from API

  # querying agency data (for ORI values to use in API calls)
agencies_url <- "https://api.usa.gov/crime/fbi/sapi/api/agencies?"

agencies <- jsonlite::fromJSON(paste(agencies_url, API_KEY, sep="API_KEY=")) %>% 
  unlist(recursive = F) %>% 
  do.call(what=rbind) %>% 
  as.data.frame()

fl_agencies <- filter(agencies, state_abbr=="FL")

#length(unique(agencies$state_abbr))    we find that there are 54 states currently, will reduce to 51 by removing PR, VI, and GM

agencies <- filter(agencies, 
                   state_abbr!="PR" &
                     state_abbr!="GM" &
                     state_abbr!="VI")
  

## delete eventually
victim_data_url_1 <- "https://api.usa.gov/crime/fbi/sapi/api/data/nibrs"
victim_data_url_2 <- "victim/agencies"
paste(paste(victim_data_url_1, "all-offenses", victim_data_url_2, "FL0500000", "race?", sep="/"), API_KEY, sep="API_KEY=")

jsonlite::fromJSON(paste(paste(victim_data_url_1, "all-offenses", victim_data_url_2, "FL0500000", "race?", sep="/"), API_KEY, sep="API_KEY="))$results
##

rm(agencies_url)
```

```{r echo=F, warning=F, message=F}
# Police Employment

#The Police Employee dataset is made up of data collected annually about law enforcement officers and civilians employed by law enforcement agencies. The dataset contains information about the number of officers and civilians employed and the rate of police employees per a location’s population.

#The Uniform Crime Reporting (UCR) Program defines law enforcement officers as individuals who ordinarily carry a firearm and a badge, have full arrest powers, and are paid from governmental funds set aside specifically for sworn law enforcement representatives.

pe <- read.csv("pe_1960_2019.csv") %>% 
  rename(state=state_postal_abbr)

pe_recent <- pe %>%
  drop_na(officer_rate_per_1000) %>% 
  filter(data_year >= 2010 & data_year <= 2019) %>% 
  filter(state != "PR")

pe_min <- pe_recent %>% 
  group_by(data_year) %>% 
  slice_min(order_by=officer_rate_per_1000, n=5)

pe_max <- pe_recent %>% 
  group_by(data_year) %>% 
  slice_max(order_by=officer_rate_per_1000, n=5)

```

```{r echo=F, warning=F, message=F}
# Consider getting API level police employment data (get male and female officer data)
```

```{r echo=F, warning=F, message=F}
# Import national level offense data (arson, rape, etc.)

# Unfortunately, state and agency level offense data is troublesome to obtain. Moreover, the FBI discourages using lower level data to make comparisons between states/agencies. As such, we will only look at national level data.

offense_url_1 <- "https://api.usa.gov/crime/fbi/sapi/api/nibrs"
offense_url_2 <- "offender/national"

offenses <- c("aggravated-assault", "burglary", "larceny", "motor-vehicle-theft", 
              "homicide", "rape", "robbery", "arson", "violent-crime", "property-crime")
cats <- c("count", "age", "race", "sex")

offenses_df <- data.frame(offense = c(),
                          stringsAsFactors=F)

for (off in offenses){
  temp <- data.frame(data_year = integer(),
                     stringsAsFactors=F)
  for (cat in cats){
    offense_url <- paste(paste(offense_url_1, off, offense_url_2, cat, sep="/"), "?", "API_KEY=", API_KEY, sep="")
    query_data <- jsonlite::fromJSON(offense_url)
    
    df <- query_data$data %>% 
      subset(select=-c(month_num)) %>% 
      group_by(data_year) %>%
      mutate(k = key) %>%
      spread(k, value) %>% 
      mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>% 
      summarise_if(is.numeric, funs(sum))
    
    colnames(df) <- colnames(df) %>% 
      lapply(tolower) %>% 
      unlist()
    if ("unknown" %in% colnames(df)){
      colnames(df)[which(colnames(df) %in% "unknown")] <- paste("unknown_", cat, sep="")
    }
    
    temp <- merge(temp, df, 
                  by="data_year",
                  all=T)
  }
  temp <- temp %>% 
    mutate(offense = off)
  offenses_df <- rbind(offenses_df, temp)
}

new_col_order <- colnames(offenses_df) %>% 
  append("offense", after=1)
new_col_order <- new_col_order[-length(new_col_order)]

offenses_df <- offenses_df[,new_col_order] %>% 
  filter(data_year >= 2010 & data_year <= 2019)

rm(offense_url_1)
rm(offense_url_2)
rm(offense_url)
rm(off)
rm(cat)
rm(query_data)
rm(df)
```

```{r echo=F, warning=F, message=F}
# Transform (i.e. merge, etc.) dataframes to get officer, civilians, type of offense, and year, with count of race, sex, age

```

```{r echo=F, warning=F, message=F}
# Create linear models
```

```{r echo=F, warning=F, message=F}
# Maps of Police Employment for each year
pe10 <- plot_usmap(data=filter(pe_recent, data_year==2010),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe11 <- plot_usmap(data=filter(pe_recent, data_year==2011),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe12 <- plot_usmap(data=filter(pe_recent, data_year==2012),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe13 <- plot_usmap(data=filter(pe_recent, data_year==2013),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe14 <- plot_usmap(data=filter(pe_recent, data_year==2014),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe15 <- plot_usmap(data=filter(pe_recent, data_year==2015),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe16 <- plot_usmap(data=filter(pe_recent, data_year==2016),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe17 <- plot_usmap(data=filter(pe_recent, data_year==2017),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe18 <- plot_usmap(data=filter(pe_recent, data_year==2018),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")

pe19 <- plot_usmap(data=filter(pe_recent, data_year==2019),
                   values="officer_rate_per_1000",
                   color="black")+
  scale_fill_continuous(low="white", high="steelblue4",
                        name="Officers per 1000")+
  theme(legend.position="none")
```

```{r echo=F, warning=F, message=F}
# Plotting 
ggarrange(pe10, pe11, pe12, pe13, pe14, pe15, pe16, pe17, pe18, pe19,
          labels=c(2010:2019),
          ncol=5, nrow=2)


```






Compare 5 highest and 5 lowest officer rate states over time.

In these states, look at which types of crimes were committed the most.






















